Davos: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-07-14 paper/main.tex
edits through section 2.2.1

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/package_structure}
\caption{\small \textbf{Package structure.} The Davos package
  comprises two interdependent subpackages.  The \texttt{davos.core}
  subpackage includes modules for parsing \texttt{smuggle} statements
  and onion comments, installing and validating packages, and
  configuring Davos's behavior.  The
  \texttt{davos.implementations} subpackage includes
  environment-specific modifications and features that are needed to
  support the core functionality across different notebook-based
  environments.  Individual modules (i.e., \texttt{.py} files) are represented by lime
  rounded rectangles, and arrows denote dependencies (each arrow
  points to a module that imports objects defined in the module at the
  arrow's source).}
\label{fig:package-structure}
\end{figure}

@@

The Davos package consists of two interdependent subpackages
(see Fig.~\ref{fig:package-structure}). The first,
\texttt{davos.core}, comprises a set of modules that
implement the bulk of the package's core
functionality, including pipelines for installing and validating
packages, custom parsers for the \texttt{smuggle} statement (see
Sec.~\ref{subsec:smuggle}) and onion comment (see
Sec.~\ref{subsec:onion}), a system for isolating dependencies of
different projects (see Sec.~\ref{subsec:projects}), and a runtime
interface for configuring Davos's behavior (see Sec.~\ref{subsec:config}).
However, certain critical aspects of this
functionality require (often substantially) different implementations
depending on properties of the notebook environment in which
Davos is used (e.g., whether the frontend is provided by
Jupyter or Google Colaboratory, or which version of
IPython~\cite{PereGran07} is used by the notebook kernel). To deal
with this, environment-dependent components of core features and behaviors
are isolated and abstracted to ``helper functions'' in the
\texttt{davos.implementations} subpackage. This second subpackage
defines multiple, interchangeable versions of each helper function,
organized into modules by the conditions that trigger their use. At
runtime, Davos detects various features in the notebook
environment and selectively imports a single version of each helper
function into the top-level \texttt{davos.implementations} namespace,
allowing \texttt{davos.core} modules to access the proper
implementations for the current notebook environment in a single,
consistent location. An additional benefit of this design is that it
allows maintainers, developers, and users to extend Davos to
support new, updated, or custom notebook variants by creating new
\texttt{davos.implementations} modules that define their own versions
of each helper function, modified from existing implementations as
needed.

@@

Functionally, importing Davos in an IPython notebook enables
an additional Python keyword: ``\texttt{smuggle}'' (see
Sec.~\ref{subsec:implementation} for details on how this works).
The \texttt{smuggle} keyword-like object can be used as a drop-in
replacement for Python's built-in \texttt{import} keyword to load
packages, modules, and other objects into the current namespace.
However, whereas \texttt{import} will fail if the requested package is
not installed locally, \texttt{smuggle} statements can handle missing
packages on the fly.  If a smuggled package does not exist in the
local environment, Davos will download and install it automatically,
expose its contents to Python's \texttt{import} machinery, and load it
into the namespace for immediate use.

Installing new packages in a notebook using standard approaches (e.g., system
commmands) affect the runtime environment. This could lead to undesired
behaviors. For example, running a notebook that installs new packages in the
user's primary system environment might alter their main system installation
and/or containerized environment in unexpected ways (e.g., changing package
versions, causing conflicts with other packages, etc.). To protect against
undesired changes to the runtime environment, Davos incorporates its
own virtual environment for managing packages it installs. When Davos
is imported, a new virtual environment (folder) is created automatically. The
folder's name may be customized to support multi-notebook projects. Any
\texttt{smuggle}d packages that were not available in the notebook's runtime
environment are installed to the current project folder. The runtime
environment remains unaffected by Davos's behavior (see
Sec.~\ref{subsec:projects}).

>>>>

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/package_structure}
\caption{\small \textbf{Package structure.} The Davos package
  comprises two interdependent subpackages.  The \texttt{davos.core}
  subpackage includes modules for parsing \texttt{smuggle} statements
  and onion comments, installing and validating packages, isolating and managing and
  configuring Davos's behavior.  The
  \texttt{davos.implementations} subpackage includes
  environment-specific modifications and features that are needed to
  support the core functionality across different notebook-based
  environments.  Individual modules (i.e., \texttt{.py} files) are represented by lime
  rounded rectangles, and arrows denote dependencies (each arrow
  points to a module that imports objects defined in the module at the
  arrow's source).}
\label{fig:package-structure}
\end{figure}

@@

The Davos package consists of two interdependent subpackages
(see Fig.~\ref{fig:package-structure}). The first,
\texttt{davos.core}, comprises a set of modules that
implement the bulk of the package's core
functionality, including pipelines for installing and validating
packages, custom parsers for the \texttt{smuggle} statement (see
Sec.~\ref{subsec:smuggle}) and onion comment (see
Sec.~\ref{subsec:onion}), a system for isolating dependencies of
different projects (see Sec.~\ref{subsec:projects}), and a runtime
interface for configuring Davos's behavior (see Sec.~\ref{subsec:config}).
However, certain critical aspects of this
functionality require (often substantially) different implementations
depending on properties of the notebook environment in which
Davos is used (e.g., whether the frontend is provided by
Jupyter or Google Colaboratory, or which version of
IPython~\cite{PereGran07} is used by the notebook kernel). To deal
with this, environment-dependent components of core features and behaviors
are isolated and abstracted to ``helper functions'' in the
\texttt{davos.implementations} subpackage. This second subpackage
defines multiple, interchangeable versions of each helper function,
organized into modules by the conditions that trigger their use. At
runtime, Davos detects various features in the notebook
environment and selectively imports a single version of each helper
function into the top-level \texttt{davos.implementations} namespace,
allowing \texttt{davos.core} modules to access the proper
implementations for the current notebook environment in a single,
consistent location. An additional benefit of this design is that it
allows both maintainers and users to easily extend Davos to
support new, updated, or custom notebook variants by adding new
\texttt{davos.implementations} modules that define their own versions
of each helper function, modified from existing implementations as
needed.

@@

Functionally, importing Davos in an IPython notebook enables
an additional Python keyword: ``\texttt{smuggle}'' (see
Sec.~\ref{subsec:implementation} for details on how this works).
The \texttt{smuggle} keyword can be used as a drop-in
replacement for Python's built-in \texttt{import} keyword to load
packages, modules, and other objects into the notebook's namespace.
However, whereas \texttt{import} will fail if the requested package is
not installed locally, \texttt{smuggle} statements can handle missing
packages on the fly.  If a smuggled package does not exist in the
user's Python environment, Davos will download and install it automatically,
expose its contents to Python's \texttt{import} machinery, and load it
into the notebook for immediate use.

Importantly, packages installed by Davos are made available for use in the
notebook without affecting the user's Python environment. By default,
\texttt{smuggle} statements install missing packages (and any missing
dependencies of those packages) into a notebook-specific, virtual
environment-like directory called a ``project'' (see
Sec.~\ref{subsec:projects}). When a \texttt{smuggle} statement in a particular
notebook is executed, Davos will first check for the smuggled package in that
notebook's corresponding project directory before searching the user's main
Python environment. In this way, \texttt{smuggle} statements can be substituted
for \texttt{import} statements to automatically ensure that all packages needed
to run a notebook are installed and available at runtime without risking
interfering with dependencies of the user's other Python programs, or other
Davos-enhanced notebooks.
